NMRlipids Databank: Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration

### 2022-10-06 main.tex
Update text

To exemplify the finding of optimal model for a certain lipid, the top ranking simulation for the order parameter quality of POPE lipid are shown in Fig.~\ref{fig:quality} B.

Differences in the predictions of area per lipids for POPC and POPE bilayers between different force fields are demonstrated in Fig.~\ref{fig:POPCPOPEapls}. Because area per lipid is strongly correlated with the average order of $\it{sn}$-1 chain ((Fig.~\ref{fig:quality} F)), its order parameters can used as a proxy for the area per lipid to evaluate its quality against experiments. Simulations with the largest area per lipids for POPE in Fig.~\ref{fig:POPCPOPEapls}, GROMOS-CKP and Slipids, have significantly higher qualities in Fig.~\ref{fig:quality} B than lipid17 and CHARMM36 with smaller are per lipids.  
Among the simulations available in the NMRlipids databank, the Slipids force field gives the best quality in terms of acyl chain order parameters and form factor, and predicts the largest area per molecule for POPE and smallest difference with POPC. In conclusion, simulations with Slipids is the most reliable force field for membrane packing in POPC:POPE mixtures althought its glycerol backbone is not accurately modelled. Similar comparisons utilizing the preliminary data from the NMRlipids databank have concluded that Slipids is relatively good model also for mixtures of charged POPC:POPS membranes, although models with better counterion binding have higher quality, but CHARMM36 is the best to study differences in headgroup conformational ensembles between different lipid types.

On the other hand, the first form factor minima is correlated stronger with membrane thickness (Fig.~\ref{fig:quality} C) than with the area per lipid (Fig.~\ref{fig:QualityCorrelationsSI}). 
Scatter plots and Pearson correlation coefficients in Fig.~\ref{fig:quality} C reveal strong correlation between membrane area per lipid and thickness.

\begin{figure}[tbp]
    \centering
    \includegraphics[width=88mm]{Figures/APL.pdf}
    \caption{Differences in area per lipids of POPC and POPE bilayers between different force fields.  
    }
    \label{fig:POPCPOPEapls}
\end{figure}

Understanding such complex picture of lipid bilayer MD simulation quality would not be possible without automatic quality evaluation of large sets of simulations enabled by the NMRlipids databank.

@@

\subsection{Water diffusion anisotropy in membrane systems}
The anisotropic water diffusion through and along of membranes is utilized in MRI imaging \cite{??} and plays a role in the drug translocation through biological material, particularly in skin \cite{hansen13,wen18,nitsche19,roberts21,??}. Dependencies between membrane properties, water permeablity, and anisotropic water diffusion have been recognized \cite{??} but consensus on how water permeability depends on membrane properties, such as area per lipid and thickness, or molecular composition, has not yet been reached despite of several available models~\cite{nitsche13,nitsche16,shinoda16,venable19,??}. MD simulations can be used to elucidate such dependencies, but systematic collection of sufficient amount of data has been challenging because few water permeation events are typically observed in a single MD simulation trajectory \cite{??}. Nevertheless, averaging over large amounts of data available in the NMRlipids databank enables to study the trends of water permeation upon changing memebrane properties.

%averaging Here we demonstrate the power of the NMRlipids databank by analysing how 
Water permeabilities calculated from all the simulations in the NMRlipids databank as a function of area per lipids, thickness, and acyl chain order are shown in  Fig.~\ref{fig:permeability} (a-c). To illustrate how permeability depends these properties, average over systems with fixed values of area per lipids, thickness, and acyl chain order are also shown in the figure. Permeability increases with increasing area per lipid and decreases with increasing membrane thickness and acyl chain order. Previous simulation and experimental studies have reported linear dependence of permeability on both area per lipid and thickness for limited set of lipid mixtures but not always for all types of lipid mixtures \cite{mathai08,frallicciardi22}. Our results suggest stronger dependence for thicknesses below 3.9\,nm and areas above 69\,Å$^2$, but the dependence may appear linear above and below these values as shown in the insets of Figs. \label{fig:permeability} a) and b). As expected, the permeability increases with increasing temperature in Figs. \label{fig:permeability} d). The Arrhenius plot gives 17\,$\pm$\,4\,k$_b$T for the average energy barrier for the water permeation. The values for water permeabilities from simulations in the databank vary between 0.3 and 322\,$\mu$m/s with the mean and median of 14\,$\mu$m/s and 8\,$\mu$m/s, respectively. These values have the same order of magnitude as the experimentally determined diffusive permeability coefficients, but are on average below the values reported for PC lipids in liquid crystalline phase,~190-330\,$\mu$m/s~\cite{jansen95}. 
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on hydration level, charged lipid fraction, cholesterol fraction or POPE fraction were not observed, although weak decrease for the latter two may be visible.

\begin{figure}[tbp]
    \centering
    \includegraphics[width=180mm]{Figures/permeation2.pdf}
    \caption{(b-c) Water permeation through lipid membranes analyzed from the databank as a function of temperature, thickness, area per lipid, and acyl chain order. 
      Inserts in c) and d) show the region where the dependence could be considered approximately linear.
      Insert in a) shows the Arrhenius plot of permeation ($\ln$ (P) vs. 1/T) that gives 17\,$\pm$\,4\,k$_b$T for the average activation energy for water permeation through lipid bilayer.
      (f-g) Diffusion anistoropy of water as a function of thickness and area per lipid.
      (h) Lateral diffusion of water as a function of hydration level. Experimental points for DMPC bilayers at 313\,K at different hydration levels are shown~\cite{rudakova04}.}
    \label{fig:permeability}
\end{figure}

To analyze the anisotropic water diffusion, we also calculated the water diffusion along the membrane. The results at different hydration levels are shown in Fig.~\ref{fig:permeability} h) together with the experimental data~\cite{rudakova04}. The experimental water diffusion increases with increasing hydration level toward the value for bulk water (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K)\cite{khakimov08}. Simulations overestimate the experimental bulk value approximately with the factor of 2 at high hydration levels which is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion. However, simulation results are closer to experiments with low hydration levels. To estimate the diffusion anistoropy, we translated the water permeation to diffusion coefficient through multilamellar stack using the Tanner equation~\cite{??}. The resulting diffusion coefficients trough the membranes are approximately five orders of magnitude slower than along the membrane which is at the upper limits of the anisotropic estimated from the experimental data~\cite{nitsche19}. Relatively high anisotropy is understandable as simulations give slightly slower permeation rates and higher lateral diffusion rates than experiments. Significant dependence of diffusion anisotropy on membrane thickness and area per lipid are observed in Figs.~\ref{fig:permeability} f) and g). The anisotropy becomes linearly stronger with decreasing area per lipid while the dependency is stronger with increasing thickness. These results can be explained by the increasing and decreasing permeabilities with thickness and are per lipid (Figs \ref{fig:permeability} c) and d), respectively, while lateral diffusion remains approximately constant (Fig. \ref{??} a) and c)).

>>>>

\subsection{Finding the best models for PC and PE lipid mixtures}
The power of the NMRlipids databank in finding the best models for a certain lipid bilayer system is exemplified here for mixtures of PE and PC lipids. Area per lipids from available POPC and POPE bilayer simulations in the databank from different force fields at 310\,K are shown in Fig.~\ref{fig:POPCPOPEapls}. Because the area per lipid strongly correlates with the average order of $\it{sn}$-1 chain (Fig.~\ref{fig:quality} F), we use order  parameter qualities to evaluate which simulations predict the best area per lipids for POPC and POPE. Among the force fields with area per lipids shown in Fig.~\ref{fig:POPCPOPEapls}, Slipids ranks 1st for POPC with a clear difference to others, and 2nd for POPE with only marginally lower quality than GROMOS-CKP (Figs.~\ref{fig:quality} A and B). The direct comparison with experiments in Fig.~\ref{fig:POPC_POPE_dataSI} in the supplementary information reveals that GROMOS-CKP and CHARMM36 predicts too ordered (too low area per lipid) bilayer for POPC, while the order parameters in Lipid17 are lower than in experiments indicating overestimated area per lipid. For POPE, order parameters in  Slipids and GROMOS-CKP are close to experiments suggesting that their are per lipids are most reasonable, while CHARMM36 and Lipid17 are too packed thereby overestimating the order parameters. 
\begin{figure}[tbp]
    \centering
    \includegraphics[width=88mm]{Figures/APL.pdf}
    \caption{Differences in area per lipids of POPC and POPE bilayers between different force fields.  
    }
    \label{fig:POPCPOPEapls}
\end{figure}

In conclusion, the quality evaluation based on the NMRlipids databank suggests that the Slipids parameters would give the most reliable description for membrane packing of POPC:POPE mixtures. Slipids was a relatively good model also for POPC:POPS mixtures in similar comparisons utilizing the preliminary version of the NMRlipids databank~\cite{antila22b}, although charged membranes are more complicated as the counterion binding affinity plays a significant role in membrane packing~\cite{antila19}. On the other hand, CHARMM36 overcomes Slipids when the focus is on conformational ensembles of headgroups  in different types of lipids because the accuracy of glycerol backbone is low in Slipids~\cite{bacle21}. Altogether, these examples demonstrate how the NMRlipids databank can be used to navigate in the complex landscape of lipid force field quality.

%On the other hand, the first form factor minima is correlated stronger with membrane thickness (Fig.~\ref{fig:quality} C) than with the area per lipid (Fig.~\ref{fig:QualityCorrelationsSI}). 
%Scatter plots and Pearson correlation coefficients in Fig.~\ref{fig:quality} C reveal strong correlation between membrane area per lipid and thickness.

%Understanding such complex picture of lipid bilayer MD simulation quality would not be possible without automatic quality evaluation of large sets of simulations enabled by the NMRlipids databank.

@@

\subsection{Water diffusion anisotropy in membrane systems}
The anisotropic water diffusion through and along of membranes plays a role in the drug translocation through biological material, particularly in skin \cite{hansen13,wen18,nitsche19,roberts21,??}, and in MRI imaging~\cite{??}. Therefore, dependence of water permeation through membranes and translational diffusion at membrane surfaces on membrane properties has been studied with different methods, yet consensus 
%on how water permeability depends on membrane properties, such as area per lipid and thickness, or molecular composition, 
has not been reached despite of several available models~\cite{nitsche13,nitsche16,shinoda16,venable19,??}. MD have been also used in these endeavours, but systematic studies are limited by the accessible amount of data as only few water permeation events are typically observed in a single MD simulation trajectory~\cite{??}. Therefore, this serves as an excellent example where the large amount of available MD simulations in the NMRlipids databank can be utilized to answer important biological questions.

To this end, we first calculated the water permeability through membranes from all the simulations in the NMRlipids databank. Resulting permeabilities are shown as a function of temperature, membrane thickness, area per lipid, and acyl chain order in Figs.~\ref{fig:permeability} B-E. The values for water permeabilities from simulations vary between 0.3 and 322\,$\mu$m/s with the mean and median of 14\,$\mu$m/s and 8\,$\mu$m/s, respectively. These values have the same order of magnitude as the experimentally determined diffusive permeability coefficients, but are on average below the values reported for PC lipids in liquid crystalline phase,~190-330\,$\mu$m/s~\cite{jansen95}. To analyze how water permeability through membranes depend on these properties on average, we calculated the average permeabilities over all systems with the value of temperature, thickness, area per lipid, or acyl chain order within a fixed range. These averaged permeabilities also shown in Figs.~\ref{fig:permeability} with black dots. As expected, the permeability increases with the temperature, giving 17\,$\pm$\,4\,k$_b$T for the average energy barrier for the water permeation from the Arrhenius plot in Fig.~\label{fig:permeability} B. Water permeability decreases with increasing membrane packing, i.e., with decreasing area per lipid and increasing thickness and acyl chain order. Linear dependence of permeability on area per lipid and thickness is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai08,frallicciardi22}. Our results suggest on average linear dependence for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,Å$^2$ (insets in Figs.~\ref{fig:permeability} C-D), yet stronger dependence for more loosely packed membranes. 
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%, although weak decrease for the latter two may be visible.

\begin{figure}[tb]
    \centering
    \includegraphics[width=180mm]{Figures/permeation2.pdf}
    \caption{A 
    {\bf B-E} Water permeation through membranes analyzed from the databank as a function of temperature, thickness, area per lipid, and acyl chain order. Values from simulations with non-zero permeation values are shown with blue dots. Averages over fixed range of x-axis values are shown with black dots. Insert in B) shows the Arrhenius plot of permeation ($\ln$ (P) vs. 1/T) that gives 17\,$\pm$\,4\,k$_b$T for the average activation energy for water permeation through lipid bilayer. Inserts in C) and D) show the region where the dependence could be considered approximately linear.
    {\bf F-G} Diffusion anistoropy of water as a function of thickness and area per lipid.
    {\bf H} Lateral diffusion of water as a function of hydration level. Experimental points for DMPC bilayers at 313\,K at different hydration levels are shown~\cite{rudakova04}.}
    \label{fig:permeability}
\end{figure}

To analyze the anisotropic diffusion of water in membrane environment, we also calculated the water translational diffusion along the membrane surface from all simulations in the NMRlipids databank. The results at different hydration levels are shown in Fig.~\ref{fig:permeability} H together with the experimental data~\cite{rudakova04}. 
%The experimental water diffusion increases with increasing hydration level toward the value for bulk water . 
%As expected, water translational diffusion increase in the diffusion is observed with the increased temperature and hydration level (Figs.~\ref{fig:permeability} H and~\ref{fig:diffusionSI} B). 
Simulation results are close to the experimental values at low hydration levels, but increase to the values of approximately two times higher than experimental bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) at high hydration levels. 
This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{??}. The water translational diffusion incereased with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Fig.~\ref{fig:diffusionSI} in the supplementary information. In order to estimate the diffusion anistoropy of water in multilamellar membrane system, we translated the water permeation coefficients through membranes to diffusion coefficients through multilamellar stack using the Tanner equation~\cite{??}. The resulting diffusion coefficients through the membranes are approximately five orders of magnitude slower than along the membrane (Figs.~\ref{fig:permeability} F and G), which is at the upper limits of the anisotropic estimated from the experimental data~\cite{nitsche19}. Relatively high anisotropy is understandable as simulations give slightly slower permeation rates and higher lateral diffusion rates than experiments. Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} F and G.
This follows from decreasing water permeability with membrane packing (Figs \ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Fig. \ref{fig:diffusionSI} A and C).
